---
title: "visualization basics"
date: today
format:
    html:
        highlight-style: github
        monofont: "Fira Code"
        code-overflow: wrap
        toc: true
        theme: 
            light: flatly
            dark: darkly
---

# Why visualization?
Data visualization  

- provides preliminary understanding of the data
- helps present and disseminate knowledge
- is a relatively under-developed subject of high-dimensional data science

Different features of data and differnt focuses of visualization require different visualization softwares

# Two visualization paradigms

- **Static visualization:** a snapshot of a dataset (non-interactive)
- **Dynamic:** a sequence of snapshots of a dataset, or a snapshot of a dataset that a user can interact with (like using the shiny package in R)

## Static visualization
There are several R packages to produce static visualizations.  

- `ggplot2` is a popularized package for static visualization. This is the library that we will be using this semester.
- R command `plot` is a basic function for static visualization
- `igraph` creates graphs (as representations of network data such as friendship networks)
  - works like python's `networkx` package
- `ggdendro` creates dendrograms and tree diagrams for clustering tasks
- `ggmpap` visualizes data frames over maps (e.g. crime data over a map of the US)

```{r}
#| code-fold: true
library(ggplot2)
#create a random scatter plot
set.seed(123)
df <- data.frame(x = rnorm(100), y = rnorm(100))

ggplot(df, aes(x, y)) + geom_point()
```

## Dynamic visualization
There are also R packages for dynamic data visualizations

- `plotly` which visualizes data frames dynamically
- `gganimate` which visualizes data frame dynamically.

These packages and those mentioned in the previous slide are somehow connected with and based on `ggplot2`.

The below plot is created using `plotly`

```{r, message=FALSE}
#| code-fold: true
library(plotly)
library(gapminder)

p <- gapminder %>%
  filter(year==1977) %>%
  ggplot( aes(gdpPercap, lifeExp, size = pop, color=continent)) +
  geom_point() +
  theme_bw()

ggplotly(p)
```

# ggplot2

The elements involved in a ggplot2 plot are:

- data, aesthetic mappings, geometric objects
- statistical transformations
- scales, coordinate system
- facets

These elements will be integrated via ggplot2 grammar  
In short:  

- a plot maps data to visual elements via specific grammar
- ggplot2 builds a plot layer by layer

## On some terms
- aesthetic mappings describe how variables in data are mapped to aesthetic attributes we can perceiv

## An illustration on `ggplot2`

The `mpg` data frame comes with `ggplot2` and will be used to illustrate the grammar of graphics.

```{r}
head(mpg)
```

Plot `displ` vs `hwy` with `geom_point()`.

```{r}
p1 <- ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))
p1
```

- `p1` is the handle for the plot
- the geometric object `geom_point()` is "point", which results in a scatter plot
- the aesthetic mapping `mapping = aes` maps `displ` and `hwy` to `x` and `y` respectively

```{r}
ggplot(mpg) + 
  geom_line(mapping = aes(x = displ, y = hwy))
```

- Here we use `geom_line` which results in a line plot

### A graphic template

The basic structure of a ggplot2 command is:  
```r
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

## Elementary Visualization
Using the `mpg` dataset we will illustrate several instances of elementary visualization.  

- scatter plot
- density plot, histogram
- box plot
- bar plot, pie chart

For this section we're only going to focus on a few variables in the `mpg` dataset.  

```{r}
mptmp <- mpg %>% select(displ, hwy, drv, class)
head(mptmp)
```

Convert character variables in `mpg` into factors by using `dplyr::mutate_if()`
  
```{r}
mpg = mpg %>% dplyr::mutate_if(is.character, as.factor)
class(mpg$class)
```

### Scatter plot

```{r}
p1
```

#### color
```{r}
#| code-fold: true
#plot displ vs hwy where each class is a different color
p1a <- ggplot(mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
p1a
```

#### shape
```{r}
#| code-fold: true
#plot displ vs hwy where each class is a different shape
p1b <- ggplot(mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
p1b
```

since there are more than 6 class levels but shape values are not manually provided. Namely, unless otherwise specified, only 6 default shapes are used

this can be resolved like so
  
```{r}
p1b <- ggplot(mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, shape = class)) +
  scale_shape_manual(values = 1:length(unique(mpg$class)))

p1b
```

#### Adding axis labels and title

```{r}
#| code-fold: true
p1b + xlab("Engine Displacement") + ylab("Highway MPG") +
  ggtitle("Efficiency") + theme(plot.title = element_text(hjust = 0.5))
```

:::{.callout-note}
Axis labels and title should all be characters or strings.

The command `theme(plot.title = element_text(hjust = 0.5))` centers the title.
:::