---
title: "HW 1"
author: "Zach Fechko (011711215)"
date: today
format: 
    html:
        highlight-style: dracula
        code-overflow: wrap
        theme: 
            light: flatly
            dark: darkly
    pdf:
        highlight-style: github
        code-overflow: wrap
        code-line-numbers: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(nycflights13)
library(plotly)
```

## Problem 1

```{r label="dataframes"}
# create data frame df from flights with the following attributes:
# - months 12, 1, 2, 6, 7, and 8
# - carriers UA, AA, and DL
# - distance greater than 700

flights_sml <- flights %>% select(month, carrier, distance, arr_delay)

#dataframe grouped by month
month.df <- flights_sml %>% 
    filter(month %in% c(12, 1, 2, 6, 7, 8) & carrier %in% c("UA", "AA", "DL") & distance > 700) %>% 
    group_by(month) 


#dataframe grouped by carrier
carrier.df <- flights_sml %>%
    filter(month %in% c(12, 1, 2, 6, 7, 8) & carrier %in% c("UA", "AA", "DL") & distance > 700) %>% 
    group_by(carrier)
```


### Problem 1.a
In a single plot, create a density plot for `arr_delay` for each of the 6 months with `color` aesthetic designated by month. Note that you need to convert `month` to a factor in order to create the plot. What can you say about the average `arr_delay` for each month?

```{r label="Problem 1.a"}
#convert month to factor
month.df$month <- as.factor(month.df$month)

# create density plot
p1a <- ggplot(month.df) +
    geom_density(mapping = aes(x = arr_delay, color = month)) +
    labs(title = "Density Plot of Arrival Delay by Month", x = " Arrival Delay", y = "Density") +
    theme(plot.title = element_text(hjust = 0.5))

p1a
```

In the density plot, we can see that the average arrival delay for each month centers around 0.

### Problem 1.b
In a single plot, create a boxplot for `arr_delay` for each of the 3 carriers. What can you say about the average `arr_delay` for each carrier?

```{r label="Problem 1.b", message=FALSE}


p1b <- ggplot(carrier.df, aes(x = carrier, y = arr_delay)) +
    geom_boxplot() + 
    stat_summary(fun.y = mean, geom = "point", shape = 18, size = 3) +
    labs(title = "Boxplot of Arrival Delay by Carrier", x = "Carrier", y = "Arrival Delay") +
    theme(plot.title = element_text(hjust = 0.5))
    

p1b
```

| Carrier | Mean Arrival Delay |
|:--------:|:-------------------:|
| AA | 2.978 |
| DL | 5.112 |
| UA | 7.599 |

United airlines had the highest average arrival delay, followed by Delta, and then American.

### Problem 1.c
Create a pie chart for the 3 carriers where the percentages are the proportions of observations and where percentages are superimposed on the sectors of the pie chart.

```{r label="Problem 1.c", message=FALSE}
library(scales)

df1c <- carrier.df %>%
    group_by(carrier) %>%
    dplyr::count() %>% ungroup() %>%
    mutate(percentage = n/sum(n)) %>%
    dplyr::arrange(desc(carrier))


df1c$labels <- scales::percent(df1c$percentage)

# create pie chart
p1c <- ggplot(df1c, aes(x = "", y = percentage, fill = carrier)) +
    geom_bar(width = 1, stat = "identity") +
    geom_text(aes(label = labels), position = position_stack(vjust = 0.5)) +
    coord_polar("y", start = 0) +
    labs(title = "Proportion of Observations by Carrier", x = "", y = "Proportion") +
    theme_void() +
    theme(plot.title = element_text(hjust = 0.5))

p1c
```

### Problem 1.d
Plot `arr_delay` against `distance` with `facet_grid` designated by `month` and `carrier`

```{r label="Problem 1.d"}
df1d <- flights_sml %>%
    filter(month %in% c(12, 1, 2, 6, 7, 8) & carrier %in% c("UA", "AA", "DL") & distance > 700)

p1d <- ggplot(df1d, aes(x = distance, y = arr_delay)) +
    geom_point() +
    facet_grid(month ~ carrier) +
    labs(title = "Arrival Delay vs Distance by Month and Carrier", x = "Distance", y = "Arrival Delay") +
    theme(plot.title = element_text(hjust = 0.5))

p1d
```


### Problem 1.e
For each feasible combination of values of `month` and `carrier`, compute the sample average of `arr_delay` and save them into the variable `mean_arr_delay`, and compute the sample average of `distance` and save these averages into the variable `mean_distance`.

Plot `month` against `mean_arr_delay` with shape designated by `carrier` and color by `mean_distance` and annotate each point by its associated carrier name.

```{r label="Problem 1.e"}
df1e <- flights_sml %>%
    filter(month %in% c(12, 1, 2, 6, 7, 8) & carrier %in% c("UA", "AA", "DL") & distance > 700) %>%
    group_by(month, carrier) %>%
    summarise(mean_arr_delay = mean(arr_delay, na.rm = TRUE), mean_distance = mean(distance))

df1e

p1e <- ggplot(df1e, aes(x = month, y = mean_arr_delay, shape = carrier, color = mean_distance)) +
    geom_point() +
    labs(title = "Mean Arrival Delay by Month and Carrier", x = "Month", y = "Mean Arrival Delay") +
    scale_color_continuous(name = "Mean Distance") +
    geom_text(aes(label = carrier), nudge_x = 0.5) +
    theme(plot.title = element_text(hjust = 0.5))

p1e
```

## Problem 2

Refer to the `mpg` dataset. Plot `displ` against `hwy` with faceting by `drv` and `cyl`, color designated by `class`, and `shape` by `trans`

```{r label="Problem 2"}

mpg_sml <- mpg %>% select(displ, hwy, drv, cyl, class, trans)


p2 <- ggplot(mpg_sml, aes(x = displ, y = hwy, color = class, shape = trans)) +
    geom_point() +
    facet_grid(drv ~ cyl) +
    labs(title = "Highway MPG vs Engine Displacement by Drive Type and Cylinders", x = "Engine Displacement", y = "Highway MPG") +
    scale_color_discrete(name = "Vehicle Class") +
    scale_shape_manual(values = 1:length(unique(mpg_sml$trans)), name = "Transmission Type")

p2
```
